---
title: "BikeShare Analysis"
author: "Casey Rodriguez"
date: '2022-09-26'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("../BikeSharing/")

rm(list = ls())
df = read.csv2("bikeSharing.csv",sep=",")

library(tidyverse)
library(lubridate)
```

## Introduction

For the purpose of this presentation we are interested in the factors that contribute
to the frequency in which bicycles are rented from a bike share program. The data
we will be using is sourced from Kaggle, though few notes are given on the origin of
said data. The full source is available at <a href="https://www.kaggle.com/datasets/shrutipandit707/bikesharing">
https://www.kaggle.com/datasets/shrutipandit707/bikesharing</a>.

We are interested in determining factors that can predict the total daily ridership
of a specific day, given the weather conditions and what day we are preparing for.

The information provided in the dataset is difficult to parse at first glance and
we will need to work on cleaning the data before we can begin analysis. A quick
preview also reveals that many of the variables are unclear, often leaving discrete
entries as numerical when explanatory text would be helpful. One seemingly incorrect
detail is the `workingday` variable, later renamed to `is_workday`, which tracks
whether or not a given day is a working day. Our assumption was that this indicates
weekend days, but that is not the case, given the inconsistency the variable has
when compared to the day of the week a specific date would be.

```{r intro, echo=FALSE}
df %>%
    head()
```

## Data Cleaning

Several modifications to the data frame would benefit readability and improve
foundation work later. Important changes include renaming the columns, unifying
the date formatting, replacing the numbers in discrete variables, and truncating
the digits for temperature, humidity, and wind speed.

```{r cleaning}
# Rename each column
df <- df %>%
    rename(id           = instant,
           date         = dteday,
           season       = season,
           year         = yr,
           month        = mnth,
           is_holiday   = holiday,
           day_of_week  = weekday,
           is_workday   = workingday,
           weather      = weathersit,
           temperature  = temp,
           ambient_temp = atemp,
           humidity     = hum,
           wind_speed   = windspeed,
           casual_rides = casual,
           return_rides = registered,
           total_rides  = cnt) %>%
    # Modify column data
    mutate(# Unify date format
           date = lubridate::parse_date_time(date, orders = c('dmy')),
           # Convert discrete numbers to strings
           month        = month.name[month],
           year         = case_when(year == 0 ~ "2018",
                                    year == 1 ~ "2019"),
           season       = case_when(season == 1 ~ "Spring",
                                    season == 2 ~ "Summer",
                                    season == 3 ~ "Autumn",
                                    season == 4 ~ "Winter"),
           weather      = case_when(weather == 1 ~ "Clear",
                                    weather == 2 ~ "Overcast",
                                    weather == 3 ~ "Storm"),
           day_of_week  = wday(date, week_start=2,label=TRUE),
           is_holiday   = as.logical(is_holiday),
           is_workday   = as.logical(is_workday),
           # Vars temperature, ambient_temp, humidity, and wind_speed are characters
           # Convert them to doubles for use as continuous variables
           temperature  = as.double(temperature),
           ambient_temp = as.double(ambient_temp),
           humidity     = as.double(humidity),
           wind_speed   = as.double(wind_speed),
           # Convert relevant columns to factors
           year         = fct_relevel(year, "2018", "2019"),
           month        = fct_relevel(month, "January", "February", "March",
                                             "April", "May", "June", "July",
                                             "August", "September", "October",
                                             "November", "December"),
           season       = fct_relevel(season, "Winter", "Spring", "Summer", "Autumn"),
           weather      = fct_relevel(weather, "Clear", "Overcast", "Storm"),
           day_of_week  = fct_relevel(day_of_week, "Mon", "Tue", "Wed", "Thu",
                                                   "Fri", "Sat", "Sun"))

# Finally, preview the new data
head(df)
```

## Column Description

```{r description, echo=FALSE}
Variables = names(df)
Description = c("A unique daily identifier",
            "The date data was collected",
            "The current season",
            "The current year",
            "The current month",
            "If the day was a holiday",
            "The day of the week",
            "If the day was a workday",
            "The weather conditions of a given day",
            "Reported daily temperature throughout the day",
            "The ambient temperature throughout the day",
            "The humidity throughout the day",
            "The wind speed throughout the day",
            "How many non-registered users rented a bike",
            "How many registered users rented a bike",
            "Count of times a bike was rented on a given day")
varframe = data.frame(Variables, Description)
knitr::kable(varframe)
```

## Data Exploration

Some quick plots of ride counts across several variables can give us a quick glance
at which variables are worth investigating as good predictors of `total_rides`.

### Categorical Data

```{r EDAp1}
df %>%
    ggplot(aes(x = season, y = total_rides, fill = season)) +
    geom_violin() +
    geom_boxplot(width=0.2, color="black", fill="white") +
    labs(x = "Season",
         y = "Daily Riders",
         title = "Distribution of Daily Ridership Between Seasons") +
    theme_bw() +
    theme(legend.position = "none")

df %>%
    ggplot(aes(x = weather, y = total_rides, fill = weather)) +
    geom_violin() +
    geom_boxplot(width=0.2, color="black", fill="white") +
    labs(x = "Weather",
         y = "Daily Riders",
         title = "Distribution of Daily Ridership By Daily Weather") +
    theme_bw() +
    theme(legend.position = "none")

df %>%
    ggplot(aes(x = day_of_week, y = total_rides, fill = day_of_week)) +
    geom_violin() +
    geom_boxplot(width=0.2, color="black", fill="white") +
    labs(x = "Day of the Week",
         y = "Daily Riders",
         title = "Distribution of Daily Ridership By Day of Week") +
    theme_bw() +
    theme(legend.position = "none")

df %>%
    ggplot(aes(x = month, y = total_rides, fill = month)) +
    geom_violin() +
    geom_boxplot(width=0.2, color="black", fill="white") +
    labs(x = "Month",
         y = "Daily Riders",
         title = "Distribution of Daily Ridership By Month") +
    theme_bw() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ds <- df %>%
    select(season, month, is_holiday, day_of_week, is_workday, weather)
sapply(ds, function(x) table(x)/nrow(ds)*100)
```

Categorical variables out of the way, the vast majority (97%) of our data points
take place on non-holiday days, making it unlikely to be a good predictor for future
total ride counts. However, we still need to explore our numerical variables.

### Continuous Data

```{r EDAp2}

df %>%
    ggplot(aes(x = temperature, y = total_rides)) +
    geom_point() + 
    geom_smooth(method = "lm", se = TRUE, color = "red") +
    labs(x = "Actual Temperature (°C)",
         y = "Daily Riders",
         title = "Daily Ridership With Respect to Temperature") +
    theme_bw()

df %>%
    ggplot(aes(x = ambient_temp, y = total_rides)) +
    geom_point() + 
    geom_smooth(method = "lm", se = TRUE, color = "red") +
    labs(x = "Ambient Temperature (°C)",
         y = "Daily Riders",
         title = "Daily Ridership With Respect to Ambient Temperature") +
    theme_bw()

df %>%
    ggplot(aes(x = humidity, y = total_rides)) +
    geom_point() + 
    geom_smooth(method = "lm", se = TRUE, color = "red") +
    labs(x = "%Humidity",
         y = "Daily Riders",
         title = "Daily Ridership With Respect to Humidity") +
    theme_bw()

df %>%
    ggplot(aes(x = wind_speed, y = total_rides)) +
    geom_point() + 
    geom_smooth(method = "lm", se = TRUE, color = "red") +
    labs(x = "Wind Speed (km/h)",
         y = "Daily Riders",
         title = "Daily Ridership With Respect to Wind Speeds") +
    theme_bw()
```

From our plotting of the daily rider total against our numerical data, we see
correlation between both actual and ambient temperatures and total daily rides.
By combining our findings from our categorical and continuous exploration, we
can start to look at building our first model.

## Making a Model

Making a model is easy. Making a good model takes time. Sometimes the simplest of
models is enough to make a sufficient prediction, taking a single variable and gauging
how well it can predict a second variable. Let's see how simple models fare at
calculating daily ride totals.

```{r modelsp1}
season_rides = lm(total_rides ~ season, data=df)
summary(season_rides)

rain_rides = lm(total_rides ~ weather, data=df)
summary(rain_rides)

month_rides = lm(total_rides ~ month, data = df)
summary(month_rides)
```

While the single, categorical models look promising, running a sanity check by
looking for a random distribution of residuals tells a different story.

```{r modelsp2}
plot(season_rides$residuals, ylab = "Ride Count ~ Season Model Residuals",
     main="Ride Count ~ Season Residual Distribution")
plot(rain_rides$residuals, ylab = "Ride Count ~ Weather Model Residuals",
     main="Ride Count ~ Weather Residual Distribution")
plot(month_rides$residuals, ylab = "Ride Count ~ Month Model Residual",
     main="Ride Count ~ Month Residual Distribution")
```

The plots of each model's residuals have clear patterns, indicating that a more
complex model is warranted. Given the jump in the middle of the chart, `year`
might be more relevant than initially thought. Additionally, each variable
tested above might have unexplored correlation. 

Before moving on to multivariate linear regression, modeling how our continuous
variables match to daily rides can give more indication on what factors correlate.

```{r modelsp3}
temp_rides = lm(total_rides ~ temperature, data = df)
summary(temp_rides)

ambient_rides = lm(total_rides ~ ambient_temp, data = df)
summary(ambient_rides)

plot(temp_rides$residuals, ylab = "Ride Count ~ Temperature Model Residual",
     main="Ride Count ~ Temperature Residual Distribution")
plot(ambient_rides$residuals, ylab = "Ride Count ~ Ambient Temperature Model Residual",
     main="Ride Count ~ Ambient Temperature Residual Distribution")
```

## Making Better Models

Where single-variable regression fails, multiple-variable regression saves the day.
By introducing a more complex model, we both decrease the intuitiveness and increase
the accuracy the model provides.

Our first attempt at a multivariate model uses ambient temperatures, the year,
and the weather to predict a day's total ride count.

```{r modelsp4}
multi_rides = lm(total_rides ~ temperature + year + weather, data = df)
summary(multi_rides)

plot(multi_rides$residuals, ylab = "Ride Count ~ Temperature + Year + Weather Residuals",
     main = "Ride Count ~ Temperature + Year + Weather Residual Distribution")
```

While the first attempt did produce a more accurate chart, nearly doubling the 
R-squared values of the previous, single-variable models, it still could be better.
In particular, the residual chart seems to indicate a regular, repeating pattern.
Including seasonality in our model improves the design once more.

## The Final Model

```{r modelsp5}

all_rides = lm(total_rides ~ temperature + year + season + weather, data = df)
summary(all_rides)
plot(all_rides$residuals, ylab = "Ride Count ~ Temp + Year + Season + Weather Residuals",
     main = "Ride Count ~ Temp + Year + Season + Weather Residual Distribution")
```

With a residual distribution that lacks a clear pattern, a p-value that is well
below the standard 0.05 required for a 95% null-hypothesis rejection certainty,
and an R-squared value at approximately 0.8, the new model incorporating the area's
temperature and weather alongside the record's year and astronomical season provides
a worthwhile prediction for the day's total ride count.

## Final Model Findings

Based on the final model, our top three predictor variables are:
```{r finaldesc, echo=FALSE}
Variable = c("Temperature", "Year", "(Stormy) Weather")
Description = c("A coefficient value of 121.391 indicates that a unit increase in temperature increases the day's ride count by 121.391 rides.",
                "A coefficient value of 2052.907 indicates that, with respect to 2018, the same day in the year 2019 will see an increase in the day's ride count by 2052.907 rides.",
                "A coefficient value of -2408.860 indicates that, with respect to a sunny day, stormy weather will see 2408.860 less daily riders.")
varframe = data.frame(Variable, Description)
knitr::kable(varframe)
```

Put in common tongue, a bike rental company can expect more customers the longer
they are in business and the warmer the day is. However, they can expect a noticeable
drop in customers should the weather forecast predict rain.

Our remaining variables are also worth considering, winter brings in the most members,
with respect to spring, summer, and autumn days. Finally, sunny weather sees the most
foot traffic, with respect to both overcast and stormy weather days. For planning
purposes, the rental company should expect high demands on sunny, warm, winter days,
especially if the company has been operating in the area for over a year. On colder,
cloudy, spring days newer bike rentals can expect lower ridership numbers.

## Further Analysis

With our model complete, we can predict total ridership for a given date should
we know the expected weather conditions in advance. However, there are more details
we can explore, leaving questions for further analysis down the line. In future
study we would like to investigate how weather conditions affect ridership between
members and casual users.




